Tool-driven Revolutions in Archaeological Science

### 2018-09-10 analysis/paper/jcaa-paper.Rmd

In this paper we investigate recent changes in the way computational archaeology is being done. Revolutions or paradigm-shifts in archaeology have been discussed to happen following theoretical discurses[@clark_1993; @härke_2002], but there has been an argument in philosophy of science whether revolutions in science are either ideas-driven or tool-driven [@galison_1997; @dyson_2000]. In this light we explore how many scientific disciplines have started to embrace methods to improve the reproducibility of their research and whether this change is part of a paradigm-shift. We focus on one set of tools for this change which is based on writing scripts in free and open source programming languages and review bibliometric evidence for this change in the sciences generally and in archaeology in particular. We evaluate how computational approaches to improving reproducibility and transparency in archaeology are mediated and transformed by digital approaches and test if the adoption of these new tools may be considered a revolution or some other process.

@@

The use of R or any other scripting languages enables a major shift in the technical and social dynamics of the research community, as it offers the possibility to share one's code. The sharing of code has been identified in other disciplines as an amplifier of change in methods[@touchon_and_mccoy_2016]. It also enables researchers to not just reproduce the work of others, but to also evaluate them according to the scientific method. Thus, the sharing of code may be at the root of an upcoming tool-driven change for archaeology, consistent with what we see desribed in the work of @Galison [-@Galison_1997]. This change does not take place revolutionary-quick and does not replace older ideas as such (as has been a critique of Kuhn's idea-driven concept of paradigm shift by others [@meltzer_1979, @trigger_2006]).

@@

Starting with the discussion of Kuhn's influential work on paradigm-shifts in science [citation needed], we offer a second explanation on how pardigm-shifts develop in science with the help of Galison [-@galison_1997], who focuses on the material culture of a science and its tools. A change in the paradigms of a discipline may also only occur, if tools to implement this change are available. The change we are aiming for is the move to a reproducible computational archaeology, that adheres to the scientific method. The tool which we have been discussing here is R, as an example for scientific scripting languages.

>>>>

In this paper we investigate recent changes in the way computational archaeology is being done. Revolutions or paradigm-shifts in archaeology have been discussed to happen following theoretical discurses[@Clark_1993; @Härke_2002], but there has been an argument in philosophy of science whether revolutions in science are either ideas-driven or tool-driven [@Galison_1997; @Dyson_2000]. In this light we explore how many scientific disciplines have started to embrace methods to improve the reproducibility of their research and whether this change is part of a paradigm-shift. We focus on one set of tools for this change which is based on writing scripts in free and open source programming languages and review bibliometric evidence for this change in the sciences generally and in archaeology in particular. We evaluate how computational approaches to improving reproducibility and transparency in archaeology are mediated and transformed by digital approaches and test if the adoption of these new tools may be considered a revolution or some other process.

@@

The use of R or any other scripting languages enables a major shift in the technical and social dynamics of the research community, as it offers the possibility to share one's code. The sharing of code has been identified in other disciplines as an amplifier of change in methods[@Touchon_McCoy_2016]. It also enables researchers to not just reproduce the work of others, but to also evaluate them according to the scientific method. Thus, the sharing of code may be at the root of an upcoming tool-driven change for archaeology, consistent with what we see desribed in the work of @Galison_1997 [-@Galison_1997]. This change does not take place revolutionary-quick and does not replace older ideas as such (as has been a critique of Kuhn's idea-driven concept of paradigm shift by others [@Meltzer_1979, @Trigger_2006]).

@@

Starting with the discussion of Kuhn's influential work on paradigm-shifts in science [citation needed], we offer a second explanation on how pardigm-shifts develop in science with the help of Galison [-@Galison_1997], who focuses on the material culture of a science and its tools. A change in the paradigms of a discipline may also only occur, if tools to implement this change are available. The change we are aiming for is the move to a reproducible computational archaeology, that adheres to the scientific method. The tool which we have been discussing here is R, as an example for scientific scripting languages.
